---
title: |
  | Appendix S2:
  | Fitting the Lester biphasic growth model with known, fixed age-at-maturity
author: |
  | Kyle Wilson$^1$ and Andrew Honsey$^2$
  | $^1$The University of Calgary
  | $^2$University of Minnesota
date: "June 23, 2017"
output:
  html_document:
    pandoc_args: [
      "--biblio", "Appendix_S2_references.bib",
      "--csl", "methods-in-ecology-and-evolution.csl"
    ]
  fig_caption: yes
  fig_height: 6
  fig_width: 7
  fontsize: 11pt
  geometry: margin=1in
  highlight: tango
  df_print: kable
  pdf_document: default
references:
- type: article-journal
  id: WilsonInReview
  author:
  - family: Wilson
    given: K. L.
  - family: Honsey
    given: A.
  - family: Moe
    given: B.
  - family: Venturelli
    given: P.
  issued: 2017
  title: 'Growing the biphasic framework: techniques and recommendations for fitting emerging growth models'
  title-short: 'Growing the biphasic framework'
  container-title: 'Methods in Ecology and Evolution'
  volume: In Review
  issue: 
  page: 
  DOI: 
  URL: 
  language: en-GB

---

```{r setup, include=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
opts_chunk$set(tidy=TRUE)
opts_chunk$set(fig.show = "hold", collapse=TRUE)
#library(devtools)
#install.packages("knitcitations")
library(knitcitations)
cleanbib()
options("citation_format" = "pandoc")
```

This appendix is in support of the main manuscript in @WilsonInReview. The citation style language (csl) used herein is the methods-in-ecology-and-evolution.csl file which can be downloaded from https://github.com/citation-style-language/styles/blob/master/methods-in-ecology-and-evolution.csl and placed in the same directory as this .rmd file.


<br>

The following code provides a template for fitting the Lester biphasic growth model to length-at-age data `r citep("10.1098/rspb.2004.2778")`. In this example, we assume that age-at-maturity has been estimated a priori, and we treat the estimate as fixed when fitting the model. NB: When using this approach, users should ensure that their estimate for age-at-maturity aligns with the Lester model T parameter (the mean age at which individuals begin to invest energy into reproduction). 


### Data generation

First, we will generate realistic length-at-age data with a known value for age-at-maturity. The age range of the fishes will be read in as a vector of integers. True somatic growth rate can be any positive number and represents the length (in mm) accumulated per year in the late-stage juvenile phase. The variable $t_1$ represents the hypothetical age at length 0 (i.e., the *x*-intercept for the juvenile phase). We use a reasonable value for the coefficient of variation in length-at-age ($cv$) of 15%, a typical observation among most fishes (see review in `r citet("10.1016/j.fishres.2016.01.006")`). The variable $T$ (or Tmat in the R code below) represents the age-at-maturity for the population of fish. The parameter $g$ represents the proportion of energy in adult phase allocated to reproduction per year. Note that $g$ must be positive and has an intrinsic maximum such that: $$ g \sim \{0,3/(T-t_1)\}. $$

In this case, we put the true $g$ value at the halfway point between $0$ and $3/(T-t_1)$. 

```{r}
Tmat <- 10 # age-at-maturity
ages <- 1:30
h <- 45 # somatic growth in millimeters per year
t1 <- -0.2 #age when size=0 for the juvenile phase
g <- 0.5*(3/(Tmat-t1)) # proportion of energy in adult phase allocated to reproduction per year
cv <- 0.15 # coefficient of variation in size-at-age
```
<br>

Next, we convert the Lester parameters to von Bertalanffy parameters, which describe the asymptotic growth of the adult phase. We then calculate the 'true' length-at-age for both phases of growth using the parameters specified above. We use an if-else statement to select which 'true' length-at-age value applies for each age, given the *a priori* estimate of age-at-maturity. We then generate a plot of the lifetime growth trajectory, including the entire trajectories for both phases and the point at which growth transitions from the first to the second phase (i.e., age-at-maturity).

```{r plot1_code, echo = TRUE, fig.show="hold",fig.cap="Somatic growth occurs in two phases: juvenile and adult. Red dashed lines indicate the age- and length-at-maturity. Dashed grey line indicates juvenile growth (including if juveniles never matured). Solid grey line indicates adult asymptotic growth (including if individuals were born mature and were investing into reproduction from hatch/birth). Black dash line indicates the composite growth trajectory of the two phases."}
linf <- 3*h/g # conversion for the VBGF L-infinity
vbk <- log(1+g/3) # conversion for the VBGF parameter kappa
t0 <- Tmat + log(1-g*(Tmat-t1)/3)/log(1+g/3) #conversion for the VBGF parameter t0

lena_phase1 <- h*(ages-t1) # length-at-age for phase 1
lena_phase2 <- linf*(1-exp(-vbk*(ages-t0))) # length-at-age for phase 2
biphasic <- ifelse(ages<Tmat,lena_phase1,lena_phase2) #if-else statement for which phase a fish is allocating surplus energy


# generate plot of growth trajectory
layout(matrix(1:1,nrow=1,ncol=2))
plot(ages,lena_phase1, ylab="Size", xlab="Age",lty=3,type="l",col="grey50",lwd=3)
lines(ages,lena_phase2,col="grey50",lwd=3)
lines(ages,biphasic,lty=2,lwd=3)
segments(x0=Tmat,x1=Tmat,y0=0,y1=h*Tmat+t1,col='red',lty=2) #plot where maturity occurs
segments(x0=0,x1=Tmat,y0=h*Tmat+t1,y1=h*Tmat+t1,col='red',lty=2) #plot where maturity occurs
points(Tmat,h*Tmat+t1,pch=21,bg="grey50",cex=1.5)

```

<br>

Next, we will generate data using `rnorm()`. We will specify how many individuals will be sampled per age bin, and we will incorporate the coefficient of variation specified above. We compile the length-at-age data into a data frame object. We set the seed for the random number generator to make our results repeatable: `set.seed(2017)`. However, those that wish to bootstrap this approach (or alter the code for some other purpose) may wish to remove this command.

```{r}
N <- 10 # how many samples per age bin
set.seed(2017)
data <- NULL
for(i in 1:max(ages))
{
  sizes <- rnorm(N,biphasic[i],biphasic[i]*cv)
  
  size_age <- cbind(rep(i,N),sizes)
  data <- rbind(data,size_age)
}
colnames(data) <- c("Age","Size") # add column names
data <- as.data.frame(data) # convert to data frame
plot(data$Age,data$Size,xlab="Age",ylab="Size") # Plot data
points(ages,biphasic,pch=20,col='red') # add 'true' growth trajectory points
```

<br>

### Likelihood function
Next, we specify the likelihood function. We will estimate four parameters: $h$, $t_1$, $g$, and the $cv$ in length-at-age. We provide conversions for the von Bertalanffy (mature) growth parameters, and we use an if-else statement to differentiate between immature and mature growth. Finally, we used a penalized likelihood to help ensure that estimates of $g$ do not venture outside of analytical bounds (see Lester et al. 2004).


```{r}
## likelihood function 

nll <- function(theta)
{
  h <- theta[1]
  t1 <- theta[2]
  g <- exp(theta[3])
  size.cv <- theta[4]
  
  ## Convert to phase 2 VBGF parameters 
  linf <- 3*h/g
  vbk <- log(1+g/3)
  t0 <- Tmat + log(1-g*(Tmat-t1)/3)/log(1+g/3)
  
  
  ## Make predictions
  pred1 <- h*(data$Age-t1) #predicted length for phase 1
  pred2 <- linf*(1-exp(-vbk*(data$Age-t0))) #predicted length for phase 2
  pred_all <- ifelse(data$Age<Tmat,pred1,pred2) #discontinuous maturity breakpoint
  ll <- dnorm(data$Size,mean=pred_all,sd=pred_all*size.cv,log=TRUE) #normal likelihood with constant variance
  if(g<0|g>(3/(Tmat-t1))){
    nll <- 1e6 # penalized likelihood when g goes past bounds given in Lester et al. 2004; g must be > 0 OR < 3/(T-t1)
  }else{
    nll <- -sum(ll) # return the negative log-likelihood 
  }
  return(nll)
}
```
<br>

### Optimization and visualization of results
To fit the model, we must first provide and compile starting values for each parameter. We then use the `optim()` function to optimize the likelihood function for our list of parameters. As mentioned above, we treat $T$ (Tmat) as fixed, and we estimate $t_1$, $h$, $g$, and $cv$ (error term). If desired, one can easily calculate Akaike's information criterion (AIC) for post-hoc model comparisons. The remainder of the code below provides guidelines for extracting and plotting results -- see comments for details.


```{r,warning=F,message=F}
## starting  values
h.hat <- 35
t1.hat <- -1
g.hat <- log(0.2)
cv.hat <- 0.1

theta <- c(h.hat,t1.hat,g.hat,cv.hat) # initial parameter estimates

## optimization
fit <- optim(theta,nll,method='Nelder-Mead',control=list(maxit=1e5,reltol=1e-10),hessian=TRUE)

## AIC calculation
AIC <- 2*length(fit$par) - 2*(-fit$value)


## plot estimates with 95% asymptotically normal confidence intervals (CI)

h_pred <- fit$par[1] # extract h estimate
t1_pred <- fit$par[2] # extract t1 estimate
g_pred <- exp(fit$par[3]) # extract and back-transform g estimate
par.hat <- c(h_pred,t1_pred,log(g_pred),fit$par[4]) # compile parameter estimates
par.true <- c(h,t1,log(g),cv) # compile 'true' parameter values
fisher_info <- solve(fit$hessian) #take the inverse of the hessian to get the variance-covariance matrix
SE.par <- sqrt(diag(fisher_info)) # square-root the variance-covariance matrix to get standard errors

UI <- par.hat+1.96*SE.par # upper bounds of 95% CIs
LI <- par.hat-1.96*SE.par # lower bounds of 95% CIs

# generate plot of percent bias in parameter estimates
plot(1:4,(par.hat-par.true)/par.true*100,ylim=range((c(LI,UI)-par.true)/par.true*100),xaxt='n',
     ylab='Percent bias',xlab="Lester model parameters")
segments(x0=1:4,x1=1:4,y0=(LI-par.true)/par.true*100,
         y1=(UI-par.true)/par.true*100,
         lty=2,col="black")
axis(1,at=1:4,labels=c("h","t1","ln(g)","cv"))
abline(h=1,lty=3,col='red')


## Plot 'true' vs. estimated growth trajectory

# make conversions to phase 2 VBGF parameters
linf.hat <- 3*h_pred/g_pred
vbk.hat <- log(1+g_pred/3)
t0.hat <- Tmat + log(1-g_pred*(Tmat-t1_pred)/3)/log(1+g_pred/3)

pred.phase1 <- h_pred*(ages-t1_pred) # predicted growth in phase 1
pred.phase2 <- linf.hat*(1-exp(-vbk.hat*(ages-t0))) # predicted growth in phase 2
pred_all <- ifelse(ages<Tmat,pred.phase1,pred.phase2) # all predictions, similar to likelihood function

# generate plot
plot(ages,biphasic,type='l',ylab="Size",xlab="Age")
lines(ages,pred_all,col='red',lty=2)

```

The top panel shows the percent bias for the four estimated parameters, compared to the 'true' values. In this case, the estimates are rather unbiased (although the confidence interval is rather large for $t_1$). The lower panel displays the estimated growth trajectory (red, dashed line) compared to the 'true' growth trajectory (black line).


## References
```{r references, echo=FALSE, message=FALSE}
write.bibtex(file="Appendix_S2_references.bib")
```
